Finally Tagless, Partially Evaluated

### 2008-11-03 tagless/tagless-final.tex
Address some of Olivier's comments and at least record the rest to deal with shortly

\begin{figure}
    \begin{floatrule}
    \begin{proofrules}
        \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
        \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
        \[ e_1:t_1\to t_2 \quad e_2:t_1 \justifies e_1 e_2: t_2 \]
        \[ \text{$n$ is an integer} \justifies n:\ZZ \]
        \[ \text{$b$ is a boolean} \justifies b:\BB \]
        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
    \end{proofrules}
    \end{floatrule}
    \caption{Our typed object language}
    \label{fig:object}
\end{figure}

@@

We attack the problem of tagless (staged) type-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.
\begin{enumerate}
\item We build the first \emph{family} of interpreters, each instantiating the
   \emph{same} signature, that evaluate (\S\ref{language}),
   compile (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item These interpreters use no type tags
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
    Yet the type system of the metalanguage
    assures statically that each object program is well-typed and closed,
    and that each interpreter preserves types and never gets stuck.
    In particular, our (online) partial evaluator and CPS transformers
    avoid GADTs in their implementation and stay portable across
    Haskell 98 and ML, by expressing in their interface an inductive
    map from input types to output types.
\item Our clean, comparable implementations using OCaml modules and Haskell
    type classes show how to parametrize our final representation of object
    terms over multiple ways to assign them meanings.
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
%% \item We describe an approach to self\hyp interpretation compatible with the
%%   above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
%%   out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\item We show how to use higher-kinded abstraction to build
    embedded DSLs.
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
and \cite{Thiemann-combinators}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Conversely, the type system of the metalanguage checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, the type checker will complain there that the expected type
|Int| does not match the inferred |Bool|.  Similarly, the object term
$\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Both Haskell's and MetaOCaml's type checkers also flag syntactically invalid
object terms, such as if we forget |app| somewhere above.

@@

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, whenever the 
argument is static, we unfold |fix| rather than residualize it.
A conservative alternative is to unfold recursion only once, then residualize:
\begin{code}
let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
\end{code}
Many sophisticated approaches have been developed to decide how much to unfold
\citep{Jones-Mix,jones-partial}, but this issue is 
orthogonal to our presentation.
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{BondorfDanvy,SwadiTahaKiselyovPasalic2006}.

@@

Our partial evaluator owes much to
\citet{Thiemann-combinators} and \citet{sumii-hybrid}, who deforested
the object term representation and expressed a partial evaluator as a collection
of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
static and dynamic terms in tandem, to combine offline and online partial
evaluation.
\citearound{'s earlier self-reducers for the untyped
$\lambda$-calculus}\Citet{JFP-Mogensen,Mogensen-SelfApplicable} also
build static and dynamic terms in tandem.  However, they
build a static term for every object term, even a bound variable,
so they move some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
and remain untyped.  
The idea of generating both static and dynamic components alongside each
other seems to be in found in an early form in \citet[\S5]{consel-schism-93}. 
In contrast, we follow Sperber, Asai,
and \citeauthor{sumii-hybrid} in leaving the static term
optional, so as to perform lifting without
juggling explicit type indices in the encoding of an object term.

@@

The simply typed $\lambda$-calculus can be interpreted even just in
itself, provided we use universal types \citep{taha-tag}.  The
ensuing tagging overhead motivated \citet{Makholm-TagElim,taha-tag} 
to propose \emph{tag
elimination}, which however does not statically guarantee that all tags
will be removed \citep{WalidICFP02}.

@@

\citet{Danvy-tagging-encoding} discuss Jones optimality at length and
apply HOAS to typed self\hyp interpretation.  However, their source
language is untyped.  Therefore, their object\hyp term encoding has
tags, and their interpreter can raise run-time errors.
Nevertheless, HOAS lets the partial
evaluator remove all the tags. In contrast, our object encoding and
interpreters do not have tags to start with and obviously cannot
raise run-time errors.

@@

We are not the first to implement a typed interpreter for a typed
language.  \Citet{laod93} use type classes to implement a metacircular
interpreter (a special kind of self\hyp interpreter) of a
typed version of the SK language, which is quite different from our
object language.  Their interpreter
appears to be tagless, but they could not have implemented a
compiler or partial evaluator in the same way, since they rely
heavily on injection\hyp projection pairs.

@@

We solve the problem of embedding a typed object language in a typed
metalanguage without using GADTs, dependent types, or a universal type.
Our family of interpreters includes an evaluator, a compiler, a partial
evaluator, and CPS transformers.  It is patent that they never get stuck,
because we represent object types as metalanguage types.  This work
makes it safer and more efficient to embed DSLs
in practical metalanguages such as Haskell and ML\@.

>>>>

\begin{figure}
    \begin{floatrule}
    \begin{proofrules}
        \[ \[ [x:t_1] \proofoverdots e:t_2 \] \justifies \fun{x}e:t_1\to t_2 \]
        \[ \[ [f:t_1\to t_2] \proofoverdots e:t_1\to t_2 \] \justifies \fix{f}e:t_1\to t_2 \]
        \[ e_0:t_1\to t \quad e_1:t_1 \justifies e_0 e_1: t \]
        \[ \text{$n$ is an integer} \justifies n:\ZZ \]
        \[ \text{$b$ is a boolean} \justifies b:\BB \]
        \[ e:\BB \quad e_1:t \quad e_2:t \justifies \cond{e}{e_1}{e_2}:t \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1+e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \times e_2:\ZZ \]
        \[ e_1:\ZZ \quad e_2:\ZZ \justifies e_1 \le e_2:\BB \]
    \end{proofrules}
    \end{floatrule}
    \caption{Our typed object language}
    \label{fig:object}
\end{figure}

@@

We attack the problem of tagless (staged) type-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.
\begin{enumerate}
\item We build the first \emph{family} of interpreters, each instantiating the
   \emph{same} signature, that evaluate (\S\ref{language}),
   compile (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item These interpreters use no type tags
    and need no advanced type-system features such as GADTs, dependent types,
    or intensional type analysis.
    Yet the type system of the metalanguage
    assures statically that each object program is well-typed and closed,
    and that each interpreter preserves types and never gets stuck.
    In particular, our (online) partial evaluator and CPS transformers
    avoid GADTs in their implementation and stay portable across
    Haskell 98 and ML, by expressing in their interface an inductive
    map from input types to output types.
\item Our clean, comparable implementations using OCaml modules and Haskell
    type classes show how to parametrize our final representation of object
    terms over multiple ways to assign them meanings.
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
%% \item We describe an approach to self\hyp interpretation compatible with the
%%   above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
%%   out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\item We show how to use higher-kinded abstraction to build
    embedded DSLs.
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
and \cite{Thiemann-combinators}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Conversely, the type system of the metalanguage checks that the
represented object term is well-typed and closed.
If we err, say replace |int 7| with |bool True| in
|testpowfix7|, the type checker will complain there that the expected type
|Int| does not match the inferred |Bool|.  Similarly, the object term
$\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Both Haskell's and MetaOCaml's type checkers also flag syntactically invalid
object terms, such as if we forget |app| somewhere above.
\ccshan{Make the point that the error messages are nice because we also
metacircularly take advantage of any niceness in the error messages of
the metalanguage implementation!}

@@

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, whenever the 
argument is static, we unfold |fix| rather than residualize it.
A conservative alternative is to unfold recursion only once, then residualize:
\begin{code}
let fix f = f (pdyn (C.fix (fun x -> abstr (f (pdyn x)))))
\end{code}
Many sophisticated approaches have been developed to decide how much to unfold
\citep{Jones-Mix,jones-partial}, but this issue is 
orthogonal to our presentation.
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{BondorfDanvy,SwadiTahaKiselyovPasalic2006}.
\ccshan{Fix the citations in the preceding sentence.}

@@

Our partial evaluator owes much to
\citet{Thiemann-combinators} and \citet{sumii-hybrid}, who deforested
the object term representation and expressed a partial evaluator as a collection
of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
static and dynamic terms in tandem, to combine offline and online partial
evaluation.
\citearound{'s earlier self-reducers for the untyped
$\lambda$-calculus}\Citet{JFP-Mogensen,Mogensen-SelfApplicable} also
build static and dynamic terms in tandem.  However, they
build a static term for every object term, even a bound variable,
so they move some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
and remain untyped.  
In contrast, we follow Sperber, Asai,
and \citeauthor{sumii-hybrid} in leaving the static term
optional, so as to perform lifting without
juggling explicit type indices in the encoding of an object term.
The idea of generating static and dynamic components alongside each
other is found in an early form in Schism \citep[\S5]{consel-schism-93}. 
\ccshan{Refer to PE tradition and (earlier, even) the category-theory
concept of \emph{gluing}.}

@@

The simply typed $\lambda$-calculus can interpret itself,
provided we use universal types \citep{taha-tag}.  The
ensuing tagging overhead motivated \citet{Makholm-TagElim,taha-tag} 
to propose \emph{tag
elimination}, which however does not statically guarantee that all tags
will be removed \citep{WalidICFP02}.

@@

\ccshan{The following should be far from the first mention of optimality
in this paper.}
\citet{Danvy-tagging-encoding} discuss Jones optimality at length and
apply HOAS to typed self\hyp interpretation.  However, their source
language is untyped.  Therefore, their object\hyp term encoding has
tags, and their interpreter can raise run-time errors.
Nevertheless, HOAS lets the partial
evaluator remove all the tags. In contrast, our object encoding and
interpreters do not have tags to start with and obviously cannot
raise run-time errors.

@@

We are not the first to implement a typed interpreter for a typed
language.  \Citet{laod93} use type classes to implement a metacircular
interpreter of a
typed version of the SK language, which is quite different from our
object language.  Their interpreter
appears to be tagless, but they could not have implemented a
compiler or partial evaluator in the same way, since they rely
heavily on injection\hyp projection pairs.

@@

We solve the problem of embedding a typed object language in a typed
metalanguage without using GADTs, dependent types, or a universal type.
Our family of interpreters includes an evaluator, a compiler, a partial
evaluator, and CPS transformers.  It is patent that they never get stuck,
because we represent object types as metalanguage types.  This work
improves the safety and reduces the overhead of embedding DSLs
in practical metalanguages such as Haskell and ML\@.
